---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```

### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
set.seed(42) # Setting seed for reproducibility
population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)
# View the first few rows
head(population)
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sample_100 <- population |> 
  sample_n(100)

sample_50 <- population |> 
  sample_n(50)

sample_10 <- population |> 
  sample_n(10)
```


### Task 4: Calculate the mean age and standard deviation of the sample
```{r}
sample_100 |> summarize(mean = mean(age), sd = sd(age))
sample_50 |> summarize(mean = mean(age), sd = sd(age))
sample_10 |> summarize(mean = mean(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
```{r}
population |> summarize(mean = mean(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The results between task 4 and 5 are pretty similar after the first run through. After running it through with different sample sizes, the 50 person sample is similar to the 100 person sample, but the 10 person sample had a much different result because it is a smaller and less representative sample size. These skills can be used to gather public opinion if done well.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
