---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Jess Daninhirsch"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

ANSWER HERE
This scatterplot is a little clearer than the last one; it is showing a negative trend between the FARMS percentage and the percentage of kids who met or exceeded math expectations, which means that the schools with the higher percentage of students in the FARMS program tend to have lower math scores than schools with lower percentages of students in FARMS.

### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

ANSWER HERE
p-value:  0.00000000000000022
Adjusted R-squared value: 0.6734
coefficients: (Intercept) 68.04878    1.74143   39.08
                FARMS Pct -0.59511    0.03685  -16.15

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

ANSWER HERE
This model is worth exploring further because it has a relatively distinct pattern that could suggest a negative relationship and potential correlation between FARMS percentage and math scores.


### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

ANSWER HERE
Fox Chapel Elementary (20876) and Broad Acres Elementary (20903) have the largest residuals. The demographics of zip code 20876 (Germantown, MD) show that the median income is $101,446, but the per capita income is $48,808. The largest racial/ethnic group is Black people at 27%, followed closely by white people at 26%. Asian and Hispanic make up a majority of the other half of the population. The demographics of zip code 20903 has a lower median income at $77,349 and a per capita income of $28,359. Over half of the population is Hispanic, and a quarter of the population is Black.

https://censusreporter.org/profiles/16000US2432025-germantown-md/
https://censusreporter.org/profiles/86000US20903-20903/ 

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

ANSWER HERE
I would want to investigate why the median incomes are so different than the per capita incomes, and how that could potentially affect the number of students in the FARMS program. Additionally, since this model has more of a clear correlation, it would be worth investigating potential factors that could affect math scores that could be a result of the FARMS program, such as if students eat before a test or if they have resources at home to be able to study or do work.
