---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mnms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.
```{r}
sample5 <- mnms |> 
  sample_n(5)

sample10 <- mnms |> 
  sample_n(10)

```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
sample5 |> summarize(mean = mean(red), sd = sd(red))
sample10 |> summarize(mean = mean(red), sd = sd(red))
```



### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
mnms |> summarize(mean = mean(red), sd = sd(red))
```



### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The population mean is nearly halfway between the sample means, but there is a decent level of variance.The sample of 5 has a mean higher than the population, and the sample of 10 has a mean lower than the population. The standard deviations follow suit, with the population about halfway between the standard devations of the other samples. And once again, the sample of 5's standard deviation is higher than that of the sample of 10.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The more sampling you do, the more precise of a result you can get. The smaller sample sizes can show a taste of what the rest of the dataset looks like, but it may not be completely accurate compared to the entire dataset. But not all of the totals ended up the same, either. Random sampling works best when you have a large sample size.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
